\bf Visual Inference for Significance and Goodness-of-Fit Testing for a Social Network Model

### 2017-10-28 ASApaper/paper-child.Rnw
add rest of setup for today.

To construct a lineup, we simulate five networks from the null model and one network from the alternative model. We chose to expose our participants to only six plots at a time in order to show the node-link diagrams in more detail and to lower their cognitive load. Typically, lineup experiments are done with sets of 20 plots at a time c.f.~\citet{loy:2015, vanderplas:2016}, but we determined that not enough structure could be shown in each plot for 20 node-link diagrams. An example of a lineup like those shown to our participants is shown on the right side of the image  in Figure~\ref{fig:shiny}.  In this lineup, model M4 is the alternative model, and model M1 is the null model. To simulate lineups from the models, we set the parameters to the values given in Table~\ref{tab:fitted values} for all parameters within the respective models, with the exception of $\beta_5$. For $\beta_5$, twice the estimated value was used. More detail on why we use twice the fitted value is provided in Section~\ref{sec:setupvd}.\footnote{If you would like to explore the kinds of lineups we use in further detail, please visit \url{https://sctyner.shinyapps.io/saom_lineup_creation/}}. To get the simulations, we used the \texttt{siena07} function in \texttt{RSiena} \citep{RSiena}.

@@

\subsection{Significance Testing}

@@

The lineup generated under this scenario is shown to independent viewers. If the viewers pick out the alternative data plot, that is evidence in favor of the null hypothesis, while picking one of the null plots is evidence against the null hypothesis. The significance tests that we perform in our experiment are for $\beta_3$ and $\beta_4$, making the alternative models M3 and M4, respectively.

\subsection{Goodness-of-Fit}
\subsection{Visual Detection}\label{sec:setupvd}

Through visual inference, we want to determine at which point an effect becomes noticeable in a SAOM. By \textit{noticeable}, we mean that the inclusion of the effect alters the appearance of networks simulated from a model. An effect becomes noticeable when a simulation from a model with a noticeable effect appears different from simulations from a model where the same affect is absent.
In model M1, with only two parameters in the objective function, we varied both the density and reciprocity parameteter values one at a time. In models M2 through M5, we vary the additional parameter, $\beta_3$ through $\beta_6$. Thus, we have six different parameters of interest to us: $\beta_1, \dots, \beta_6$. We want to determine how the size of these parameters affects the overall structure of the network data simulated from the models M1 through M5, so we also vary the value of the parameters in two directions: positive and negative.

@@

<<hypothesis, fig.height=2, fig.cap="We hypothesize that as the parameter value of interest increases in absolute value, more viewers of the lineup will pick the alternative data out of a lineup.">>=
x <- seq(-10, 10, .05)
N <- length(x)
qplot(x = 1:N, y = pt(x, 1), geom = 'line') + geom_line(aes(x = -(1:N), pt(x,1))) + 
    scale_x_continuous(name = "Parameter Value", labels = c("very negative", "negative", "0", "positive", "very positive")) + 
    scale_y_continuous(name = "% detecting plot from\nthe alternative model", labels = paste0(c(0,25,50,75,100), "%"))
@

@@

In order to determine which specific values of the parameters we wanted to examine, we constructed an online application that created the lineup protocol for us to be the guinea pigs of our own experiment \citep{shiny}. A screen grab of the app is shown in Figure~\ref{fig:shiny}. On the left side of the screen, the user\footnote{Please visit \url{https://sctyner.shinyapps.io/saom_lineup_creation/} to create lineups for yourself.} can input the information necessary for creating a lineup of the models M1 through M6 for the data in Section~\ref{sec:senatedata}: first, choose to simulate only one plot from the specified model (analogous to changing the alternative model in the lineup protocol) or to simulated $M-1$ plots from the specified model (analogous to changing the null model in the lineup protocol); the model of interest; the wave of the data to examine; if model M1 is selected, whether to alter the density or the reciprocity parameter; the size of the lineup; the amount by which to multiply the effect selected; a random seed for replicability; and a layout algorithm to use for the node-link diagrams. The plot(s) that appear(s) that are \textit{not} from the model specified using the other options are simulations from model M1 with the estimates of the rate parameters, $\beta_1$ and $\beta_2$ as calculated by the procedure described in Section~\ref{sec:models}.

Using the ``Picking Lineups" web application, we settled on six parameter values to test for each of our six effects, $\beta_1, \dots \beta_6$. The complete details of the parameters tested using the lineup protocol is given in Table~\ref{tab:experdetail}. In the case of both $\beta_4$ and $\beta_6$, we could not determine a value for negative effects that made the data simulated from M3 and M5 look different than null model simulations from model M1. Because we could not detect these effects at these values, we decided that the less experienced Turkers would also not be able to. So instead of testing the negative values of these effects, we examined a different scenario: we placed 5 simulations from positive values of the parameter with one simulation from model M1 in a lineup. We did this to determine if the perception of the effect size would be symmetric: if an effect is noticed $x$\% of the time at value $\beta = 5$ when a simulation from the model is placed among null plots, then when five simulations from the model are put together with one simulation from model M1, the plot from the simpler model should be noticed about $x$\% of the time as well.

@@

We recruited 250 participants for our experiment through Amazon Mechanical Turk. Each participant read over some brief training material before beginning the experiment. After reading this material and agreeing to participate, the participants saw two trial plots, one where they were required to select the plot that was most different from the others due to its complex structure, and one where they were required to select the plot that was most different from the others due to its very simple structure. Upon getting the trial plots correct, they moved into the experiment. Each participant was randomly assigned 13 lineups to look at. The were asked to select one or more plots that they perceived as ``most different" from the others, and provide a reasoning for their choice. They could select from ``Most simple overall structure," ``Most complex overall structure," or ``Other" and provide their own text description of their reasoning. One of the 13 lineups the participants saw was the true data from the 112th senate shown in Section~\ref{sec:senatedata}. Each participant only saw the data one in order to avoid bias.

@@

For a SAOM, there are two ways a significance test of the parameters can be performed. In \texttt{RSiena}, there are $t$-type tests and Wald-type test for a single parameter and for multiple parameters. The $t$-type test statistic is simply the parameter estimate divided by its standard error, and compared to a standard normal distribution. The Wald-type test statistic for a single parameter, $\beta_k$ is 
\begin{equation}\label{eq:wald1}
\frac{(\hat{\beta_k})^2}{var(\hat{\beta_k})} \sim \chi^2_1,
\end{equation}
which is compared to a Chi-square distribution with one degree of freedom. Testing the significance of multiple parameters depends on the hypothesis we wish to test, and a $P \times K$ matrix, $A$, must be appropriately designed to test the $P$ hypotheses of interest. The null hypothesis is that $A\boldsymbol{\beta} = \mathbf{0}$, and the test statistic is    
\begin{equation}\label{eq:wald2}
(A\boldsymbol{\hat{\beta}})' \hat{\Sigma}^{-1} A\boldsymbol{\hat{\beta}} \sim \chi^2_p,
\end{equation}
where $\hat{\Sigma}$ is the estimated covariance matrix of $\boldsymbol{\beta}$. This statistic is then compared to a Chi-square distribution with $P$ degrees of freedom.

>>>>

To construct a lineup, we simulate five networks from the null model and one network from the alternative model. We chose to expose our participants to only six plots at a time in order to show the node-link diagrams in more detail and to lower their cognitive load. Typically, lineup experiments are done with sets of 20 plots at a time c.f.~\citet{loy:2015, vanderplas:2016}, but we determined that not enough structure could be shown in each plot for 20 node-link diagrams. An example of a lineup like those shown to our participants is shown on the right side of the image  in Figure~\ref{fig:shiny}.  In this lineup, model M4 is the alternative model, and model M1 is the null model. To simulate lineups from the models, we set the parameters to the values given in Table~\ref{tab:fittedvalues} for all parameters within the respective models, with the exception of $\beta_5$. For $\beta_5$, twice the estimated value was used. More detail on why we use twice the fitted value is provided in Section~\ref{sec:setupvd}.\footnote{If you would like to explore the kinds of lineups we use in further detail, please visit \url{https://sctyner.shinyapps.io/saom_lineup_creation/}}. To get the simulations, we used the \texttt{siena07} function in \texttt{RSiena} \citep{RSiena}.

@@

\subsection{Significance Testing}\label{sec:setupst}

@@

The lineup generated under this scenario is shown to independent viewers. If the viewers pick out the alternative data plot, that is evidence in favor of the null hypothesis, while picking one of the null plots is evidence against the null hypothesis. The significance tests that we perform in our experiment are for $\beta_3$ and $\beta_4$, making the alternative models M2 and M3, respectively.

\subsection{Goodness-of-Fit}

For the goodness-of-fit tests, we compare one model of interest, say M\textit{i} to the data. The hypothesis we use to generate lineups are 
\begin{description}
\item $H_0$: The data come from model M\textit{i}
\item $H_A$: The data come from some other, unknown model
\end{description}

To generate the null plots, we simulate five networks from model M\textit{i} using the corresponding parameters in Table~\ref{tab:fittedvalues}. We pick a wave to focus on, wave two, which is the first simulated network, and among these five plots, we place the true second wave data. We cannot show the data more than once to each participant, so we examine several different models in our Amazon Mechanical Turk experiment, each participant never seeing the true data wave twice. The models we chose for goodness-of-fit testing are M2, M3, M4, and M6.

\subsection{Visual Detection}\label{sec:setupvd}

Visual detection does not fit into the typical statistical testing framework as easily as the significance and goodness-of-fit tests we perform do. Through visual inference, we want to determine at which point an effect becomes noticeable in a SAOM. By \textit{noticeable}, we mean that the inclusion of the effect alters the appearance of networks simulated from a model. An effect becomes noticeable when a simulation from a model with a noticeable effect appears different from simulations from a model where the same affect is absent when they simulations are placed in a lineup. We explore all parameters in the objective function, $\beta_1, \dots, \beta_6$ in this way.

In model M1, with only two parameters in the objective function, we varied both the density and reciprocity parameteter values one at a time. In models M2 through M5, we vary the additional parameter, $\beta_3$ through $\beta_6$. Thus, we have six different parameters of interest to us: $\beta_1, \dots, \beta_6$. We want to determine how the size of these parameters affects the overall structure of the network data simulated from the models M1 through M5, so we also vary the value of the parameters in two directions: positive and negative.

@@

<<hypothesis, fig.height=2, fig.cap="We hypothesize that as the parameter value of interest increases in absolute value, more viewers of the lineup will pick the alternative data out of a lineup. Note that the significance test we construct in Section~\\ref{sec:setupst} is just one point on the line below, represented by the vertical red line.">>=
x <- seq(-10, 10, .05)
N <- length(x)
qplot(x = 1:N, y = pt(x, 1), geom = 'line') + geom_line(aes(x = -(1:N), pt(x,1))) + 
    scale_x_continuous(name = "Parameter Value", labels = c("very negative", "negative", "0", "positive", "very positive")) + 
  #geom_hline(yintercept = 1/6) +
    geom_vline(xintercept = 150, color = 'red', linetype = 'dotdash') + 
    scale_y_continuous(name = "% detecting plot from\nthe alternative model", labels = paste0(c(0,25,50,75,100), "%"))
@

@@

In order to determine which specific values of the parameters we wanted to examine, we constructed an online application that created the lineup protocol for us to be the guinea pigs of our own experiment \citep{shiny}. A screen grab of the app is shown in Figure~\ref{fig:shiny}. On the left side of the screen, the user\footnote{Please visit \url{https://sctyner.shinyapps.io/saom_lineup_creation/} to create lineups constructed from the models we present for this data for yourself.} can input the information necessary for creating a lineup of the models M1 through M6 for the data in Section~\ref{sec:senatedata}: first, choose to simulate only one plot from the specified model (analogous to changing the alternative model in the lineup protocol) or to simulated $M-1$ plots from the specified model (analogous to changing the null model in the lineup protocol); the model of interest; the wave of the data to examine; if model M1 is selected, whether to alter the density or the reciprocity parameter; the size of the lineup; the amount by which to multiply the effect selected; a random seed for replicability; and a layout algorithm to use for the node-link diagrams. There is also a checkbox if the user wishes the nodes to be colored by the size of the connected component to which they belong. The plots that appear that are \textit{not} from the model specified using the other options are simulations from model M1 with the estimates of the rate parameters, $\beta_1$ and $\beta_2$ given in Table~\ref{tab:fittedvalues}.

Using the ``Picking Lineups" web application, we settled on six parameter values to test for each of our six effects, $\beta_1, \dots \beta_6$. The complete details of the parameters tested using the lineup protocol is given in Table~\ref{tab:experdetail}. In the case of both $\beta_4$ and $\beta_6$, we could not determine a value for negative effects that made the data simulated from M3 and M5 look different than null model simulations from model M1. Because we could not detect these effects at these values, we decided that the less experienced participants in our experiment would also not be able to. So instead of testing the negative values of these effects, we examined a different scenario: we placed 5 simulations from positive values of the parameter with one simulation from model M1 in a lineup. We did this to determine if the perception of the effect size would be symmetric: if an effect is noticed $x$\% of the time at value $\beta_k = \beta_{k_0}$ when one simulation from the corresponding model is placed among five null plots from model M1, then when five simulations from the model with $\beta_k = \beta_{k_0}$ are put in a lineup with one simulation from model M1, the plot from the simpler model should be noticed about $x$\% of the time as well.

@@

We recruited 250 participants for our experiment through Amazon Mechanical Turk. Each participant read over some brief training material before beginning the experiment. After reading this material and agreeing to participate, the participants saw two trial plots, one where they were required to select the plot that was most different from the others due to its complex structure, and one where they were required to select the plot that was most different from the others due to its very simple structure. Upon getting the trial plots correct, they moved into the experiment. Each participant was randomly assigned 13 lineups to look at. The were asked to select one or more plots that they perceived as ``most different" from the others, and provide a reasoning for their choice. They could select from ``Most simple overall structure," ``Most complex overall structure," or ``Other" and provide their own text description of their reasoning. One of the 13 lineups the participants saw was the true data from the 112th senate shown in Section~\ref{sec:senatedata}. Each participant only saw the data one time in order to avoid bias. Upon completion of the 13 lineups, each participant was paid \$1.75.

@@

For a SAOM, there are two ways a conventional significance test of the parameters can be performed. In \texttt{RSiena}, there are $t$-type tests and Wald-type test for a single parameter and for multiple parameters. The $t$-type test statistic is simply the parameter estimate divided by its standard error, and compared to a standard normal distribution. The Wald-type test statistic for a single parameter, $\beta_k$ is 
\begin{equation}\label{eq:wald1}
\frac{(\hat{\beta_k})^2}{var(\hat{\beta_k})} \sim \chi^2_1,
\end{equation}
which is compared to a Chi-square distribution with one degree of freedom. Testing the significance of multiple parameters depends on the hypothesis we wish to test, and a $P \times K$ matrix, $A$, must be appropriately designed to test the $P$ hypotheses of interest. The null hypothesis is that $A\boldsymbol{\beta} = \mathbf{0}$, and the test statistic is    
\begin{equation}\label{eq:wald2}
(A\boldsymbol{\hat{\beta}})' \hat{\Sigma}^{-1} A\boldsymbol{\hat{\beta}} \sim \chi^2_p,
\end{equation}
where $\hat{\Sigma}$ is the estimated covariance matrix of $\boldsymbol{\beta}$. This statistic is then compared to a Chi-square distribution with $P$ degrees of freedom.
